Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-13 Manuscript/Manuscript.Rmd
Clean up manuscript file

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
      # | Julius Center for Health Sciences and Primary Care, 
      # | University Medical Center Utrecht, Utrecht, The Netherlands
    #affiliation2:
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
---

@@

@@

@@

<!-- -	What kinds of missingness are there? ADD: missingness mechanisms here. See e.g. @yuce08 and @hox15. -->

<!-- - Why are standard (ad hoc) missing data methods not well suited?  -->

@@

<!-- - Additional difficulty that is addressed in this tutorial: MNAR data. -->

@@

# Workflows

<!-- We'll use the IMPACT data (`metamisc::impact`) and a MAR/MNAR version of the `mice::popmis` data (i.e., a variation on the Hox (2010) popularity data, where the missingness in the variables is either missing at random (MAR) or missing not at random (MNAR)). -> ask whether we can use the Heckman repo data or simulate data ourselves -->

<!-- Heckman options: -->

<!-- - leiden85 -->

<!-- - GJRM::hiv (https://rdrr.io/github/egeminiani/GJRM/man/hiv.html) -->

<!-- - simulating -->

<!-- - IMPACT -->

@@

## Case Study I: Popularity

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
pop <- readRDS("../Data/popNCR.RDS") %>% mutate(school = round(0.5 + as.numeric(class)/10))

# ICCs
icc_pop <- icc(popular ~ as.factor(class), data = pop)
icc_teach <- icc(popteach ~ as.factor(class), data = pop)
```

`popNCR` is a simulated dataset with pupils clustered in classes, $n_{\text{participants}} = 2000$, $n_{\text{clusters}} = 100$, on 7 variables:

@@

```{r pop_pat, echo=FALSE, fig.height=6, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

@@

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# distributions
# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice

plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))

# x = "popular"
# z = "popteach"
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
  geom_density() +
  scale_color_manual(values = plot_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}

plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: pooteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal

# Feedback: 
# you only see the extreme ones in the positive sense, especially with 100 clusters
# alternatively add a quartile line around the density with geom_ribbon

# marginal

@@

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
```

@@

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
```

@@

# compare ICCs before and after imputation
ICCs <- ICCs %>% cbind(
           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
                        icc(popteach ~ class, complete(imp_predictor)), 
                        icc(texp ~ class, complete(imp_predictor)))
           )
ICCs
```

@@

@@

`impact` is traumatic brain injury data with patients clustered in studies, $n_{\text{participants}} = 11022$ and $n_{\text{clusters}} = 15$, on the following 11 variables:

@@

# TODO: check md pattern in clusters, should differ, otherwise create new missingness
# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
# sim.multi from psych
```
<!-- -> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf),  -->

<!-- TODO: also make MNAR missingness to heckman model. Maybe based on `ct` variable? Inclusion-selection variable. -> otherwise: use `leiden85` data on blood pressure with MNAR. Then run cox regression like the boshuizen article but with living situation as clusters. -> TODO: get analyses from https://www.gerkovink.com/mimp/Contents/Exercises/Day%203%20-%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html. -->

<!-- ## Amputation -->

<!-- -  -->

<!-- ## Modeling choices -->

<!-- - Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.  -->

<!-- - Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well. Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeneity. ->  -->

<!-- - Step 0: As predictor + CCA to scare off users -->

<!-- - Step 1: Random intercepts -->

<!-- - Step 2: Random slopes -->

<!-- - Step 3: Residuals -->

<!-- - Heckman model for MNAR -->

<!-- - What do the different implementations look like? How to define the imputation model(s) in \pkg{mice}? -->

<!-- ## Step 0 -->

<!-- - AKA multilevel imputation for dummies.  -->

<!-- - Doesn't work for systematic missingness.  -->

<!-- ## Step 1-3 + MNAR -->

<!-- - TODO: fill in. -->

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

# seed
set.seed(123)

@@

$$
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
                          &x_{21} &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
$$

@@

@@

@@

@@

# Case Study I: Popularity

```{r pop, echo=FALSE, message=FALSE, warning=FALSE}
# load data and merge classes into schools for this tutorial -> but this cannot work because how texp and popteach are defined...
pop <- readRDS("../Data/popNCR2.RDS") %>% mutate(class = as.factor(class))

# ICCs
icc_pop <- icc(popular ~ class, data = pop)
icc_teach <- icc(popteach ~ class, data = pop)
```

`popNCR` is a simulated dataset with pupils clustered in classes, where the number of cases $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

@@

```{r pop_pat, echo=FALSE, fig.height=4, fig.width=4, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(pop)
# TODO: add pattern by cluster, text mining example square size
# TODO: try to make it a shepley plot 
```

@@

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# plotting colors
plot_col <- mice:::mdc(c("obs", "mis")) %>% setNames(c("observed", "missing"))

# distributions
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = factor(is.na(get(z)), labels = c("missing", "observed")))) +
  geom_density(data = data %>% filter(!is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[2], color = NA) +
  geom_density(data = data %>% filter(is.na(get(z))), aes(x = get(x), group = get(cluster)), alpha = 0.01, fill = plot_col[1], color = NA) +
  geom_density() +
  scale_color_manual(values = plot_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}

# plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: legend: popteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal

# Feedback: 
# you only see the extreme ones in the positive sense, especially with 100 clusters
# alternatively add a quartile line around the density with geom_ribbon
```

@@

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
```

@@

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)

@@

# compare ICCs before and after imputation
ICCs <- ICCs %>% mutate(
           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
                        icc(popteach ~ class, complete(imp_predictor)), 
                        icc(texp ~ class, complete(imp_predictor)))
           )
ICCs
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

```{r pop_norm_eval}
# plot(imp_norm)
densityplot(imp_norm, ~popular)
ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
                    icc(popteach ~ as.factor(class), complete(imp_norm)), 
                    icc(texp ~ as.factor(class), complete(imp_norm)))
           )
ICCs
```

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan <-
  mice(
    pop %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

```{r pop_pan_eval}
# plot(imp_pan)
densityplot(imp_pan, ~popular)
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
                    icc(popteach ~ as.factor(class), complete(imp_pan)), 
                    icc(texp ~ as.factor(class), complete(imp_pan)))
           )
ICCs
```

@@

`impact` is traumatic brain injury data with patients, $n = 11022$, clustered in studies, $N = 15$. With the following 11 variables:

@@

# TODO: check md pattern in clusters, should differ, otherwise create new missingness
# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
# sim.multi from psych
```

### 2021-12-13 Manuscript/Manuscript.tex
Clean up manuscript file

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

@@

@@

@@

\hypertarget{workflows}{%
\section{Workflows}\label{workflows}}

@@

\hypertarget{case-study-i-popularity}{%
\subsection{Case Study I: Popularity}\label{case-study-i-popularity}}

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
variables:

@@

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
0.31. It would thus be wise to use multilevel modeling.

@@

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-1}
\end{figure}
\begin{figure}

{\centering \includegraphics{Manuscript_files/figure-latex/pop_dist-2}

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist-2}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2958571
2 popteach  0.3138658 0.2561198
3     texp  1.0000000 0.4696024
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2958571 0.3581435
2 popteach  0.3138658 0.2561198 0.3446899
3     texp  1.0000000 0.4696024 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\texttt{impact} is traumatic brain injury data with patients clustered
in studies, \(n_{\text{participants}} = 11022\) and
\(n_{\text{clusters}} = 15\), on the following 11 variables:

>>>>

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

\usepackage{graphicx}
\usepackage{mathtools}

@@

\[
\begin{matrix} 
                          &X1     & X2          & X3            \\
   \hline
   \mathrm{cluster 1}\{   &x_{11} &x_{12}       &\mathtt{NA}    \\
                          &x_{21} &x_{22}       &x_{32}         \\
   \mathrm{cluster 2}\{   &x_{31} &\mathtt{NA}  &x_{33}         \\
                          &x_{41} &\mathtt{NA}  &\mathtt{NA}    \\
   \mathrm{cluster 3}\{   &x_{51} &x_{52}       &x_{53}         \\
\end{matrix} 
\]

@@

@@

\hypertarget{case-study-i-popularity}{%
\section{Case Study I: Popularity}\label{case-study-i-popularity}}

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
where the number of cases \(n = 2000\), and the number of clusters
\(N = 100\), on 7 variables:

@@

The popularity data is created such that there are strong relations
between the incomplete variables and the clustering variable
\texttt{class}. We can express this using the intra-class correlation
(ICC). For \texttt{popular} the ICC is 0.33. For \texttt{popteach} it is
0.34. It would thus be wise to use multilevel modeling.

@@

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.3338626
2 popteach  0.3414766 0.3414766
3     texp  1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.3338626 0.3726455
2 popteach  0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth <- c("", "", "", "", "", "2l.norm", "")
R> imp_norm <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_norm)
R> densityplot(imp_norm, ~popular)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_norm_eval-1} \end{center}

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            norm = c(icc(popular ~ as.factor(class), complete(imp_norm)), 
+                     icc(popteach ~ as.factor(class), complete(imp_norm)), 
+                     icc(texp ~ as.factor(class), complete(imp_norm)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm
1  popular  0.3280070 0.3338626 0.3726455 0.3667167
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
R> meth <- c("", "", "", "", "", "2l.pan", "")
R> imp_pan <-
+   mice(
+     pop %>% mutate(class = as.integer(class)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # plot(imp_pan)
R> densityplot(imp_pan, ~popular)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop_pan_eval-1} \end{center}

\begin{CodeInput}
R> ICCs <- ICCs %>% mutate(
+            pan = c(icc(popular ~ as.factor(class), complete(imp_pan)), 
+                     icc(popteach ~ as.factor(class), complete(imp_pan)), 
+                     icc(texp ~ as.factor(class), complete(imp_pan)))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor      norm       pan
1  popular  0.3280070 0.3338626 0.3726455 0.3667167 0.3719078
2 popteach  0.3414766 0.3414766 0.3414766 0.3414766 0.3414766
3     texp  1.0000000 1.0000000 1.0000000 1.0000000 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\texttt{impact} is traumatic brain injury data with patients,
\(n = 11022\), clustered in studies, \(N = 15\). With the following 11
variables:
